Anomalous waves triggered by abrupt depth changes: experimental measurements and deterministic-statistical truncated KdV theory

### 2019-10-07 paper2_JNLS/paper.tex
NM

Rogue waves are abnormally large surface waves, defined by oceanographers as those that exceed twice the significant wave height \citep{muller2005rogue, ying2011linear}. Though such waves were once dismissed as myth, they have now been recorded in oceans across the globe and pose a recognized threat to seagoing vessels and naval structures. Rogue waves, also variously known as freak or anomalous waves, have been observed both in deep and shallow water, and can be triggered by a variety of mechanisms, including anomalous wind-forcing \citep{kharif2008influence, toffoli2017wind}, 
focusing due to variable bathymetry \citep{heller2008refraction, white1998chance}, 
opposing currents \citep{garrett2009rogue, onorato2011triggering},
and the Benjamin-Feir deep-water modulational instability \citep{benjamin1967disintegration, viotti2013emergence, cousins2015unsteady, farazmand2017reduced}.
The common tie between these mechanisms is their ability to generate non-normal statistics in the surface displacement; when governed by Gaussian statistics, the likelihood of a rogue wave is extremely low, but these rare events occur much more frequently when surface statistics deviate from Gaussian. In this way, it makes sense to approach anomalous waves from the broader perspective of turbulent dynamical systems \cite{sapsis2013a, sapsis2013b, sapsis2013blending, chen2016filtering, majda2016introduction, macedo2017universality, MajdaQiSIAM2018, blonigan2019extreme, guth2019machine}.

@@

Our particular focus is the laboratory experiments of Bolles et al.~\cite{bolles2019}, who demonstrated the emergence of anomalous statistics from a randomized wave-field encountering an abrupt depth change (ADC). In these experiments, the incoming wave-field is generated with nearly Gaussian statistics, and a plexiglass step placed near the middle of the tank creates the depth change. Upon passing into the shallower region, the wave distribution skews strongly towards positive displacement, with deviation from Gaussian being most pronounced a short distance downstream of the ADC. Inspired by these experimental observations, Majda et al.~\cite{majda2019} developed a theoretical framework that accurately captures several key aspects of the anomalous behavior. The theory is based on deterministic and statistical analysis of the truncated Korteweg–de Vries (TKdV) equations, and uses a combination of computational, statistical, and analytical tools. In follow-up work, Majda and Qi \cite{majdaqi2019} analyzed more severely truncated systems of lower dimension that were shown to exhibit many of the same anomalous properties while enjoying a more tractable structure.

The purpose of the present manuscript is to provide a more comprehensive treatment of the {\em combined} experimental and theoretical efforts of Bolles et al~\cite{bolles2019} and Majda et al.~\cite{majda2019, majdaqi2019}. Special emphasis is placed on the synergy between experiment and theory -- a cooperative strategy with a proven record of success \cite{camassa2012stratified, ristroph2012, ganedi2018equilibrium}. We present an assortment of previously unpublished experimental data, for example on surface slope statistics and autocorrelation, which motivates further examination and refinement of the theory. Likewise, novel theoretical predictions have inspired us to perform additional experimental measurements, particularly regarding a peculiar relationship between surface-displacement skewness and surface-slope variance that is predicted by the theory. Comparison with the new measurements further confirms the predictive power of the framework developed by Majda et al.~\cite{majda2019}. In addition, we provide a more thorough treatment of the link between experimental and model parameters to facilitate comparison between the two. In particular, we flesh out details of non-dimensionalization, which were only briefly addressed in prior work for the sake of brevity.

@@

The pivoting motion of the paddle is driven by a 5-phase stepper motor. To generate a randomized incoming wave field, the paddle angle $\phi$ is specified by a precomputed psuedo-random signal
\begin{align}
\label{PaddleAngle}
& \phi(t) = \phi_0 + \Dphi \sum_{n=1}^N a_n \cos(\omega_n t+\delta_n) \, , \\
\label{anEq}
& a_n = \sqrt{\frac{2 \Delta \omega}{\pi^{1/2} \omsig}} \, 
\exp \left( -\frac{(\omega_n - \omavg)^2}{2 \omsig^2} \right) \, ,
\end{align}
The angular frequencies are evenly spaced $\omega_n = n  \Delta \omega$ with step size $ \Delta \omega = (\omavg+4 \omsig)/N$, where $\omavg$ and $\omsig$ represent the mean and the bandwidth of $\omega$ respectively. As in prior work, all experiments reported here use the values $\omavg = \omsig = 12.5$ rad/s, corresponding to a peak forcing frequency of 2 Hz and bandwidth of 2 Hz. Meanwhile, the phases $\delta_n$ are uniformly distributed random variables. The standard deviation of the paddle angle, $\Dphi$, controls the amplitude of the generated waves. In a single experiment $\Dphi$ is fixed, and we will present a series of experiments with systematically varied values of $\Dphi$.

@@

	The free surface is illuminated by light-emitting diodes running along the bottom of the tank and is imaged from the sideview with a Nikon D3300 at 60 frames per second. The illumination technique, coupled with high pixel count of the camera, allows surface displacements to be resolved with accuracy better than 1/3 millimeter. Furthermore, these optical measurements permit extraction of wave statistics {\em continuously} in space, rather than at a few discrete locations, which is crucial for identifying regions of anomalous wave activity. Further details of the experimental setup can be found in \cite{bolles2019}.

	Example measurements of free-surface displacements $\eta$ are shown in Figs.~\ref{ExpDiagStats}(b)--(c). These measurements are extracted from the images at two representative locations: one a short distance (9 cm) upstream of the ADC and the other (15 cm) downstream of the ADC. Both signals exhibit a combination of periodic behavior and random fluctuations, with the dominant oscillations corresponding to the peak driving frequency of 2 Hz. The nature of the random fluctuations is revealed by the corresponding histograms shown in Fig.~\ref{ExpDiagStats}(d)--(e). The upstream measurements are symmetrically distributed about the mean, $\eta = 0$. In fact, Bolles et al.~found that these measurements follow a Gaussian distribution closely \cite{bolles2019}. The downstream measurements, however, are not symmetrically distributed and skew strongly towards positive displacement, $\eta > 0$. Bolles et al.~found that these measurements can be well described by a mean-zero gamma distribution \cite{bolles2019}. The slower decay rate of this downstream distribution indicates an elevated level of extreme surface displacement, i.e.~rogue waves. Bolles et al.~estimated that a rogue wave can be up to 65 times more likely in these experiments than if the displacements were Gaussian.

	The measurements shown in Fig.~\ref{ExpDiagStats} correspond to a paddle driving amplitude of $\Dphi = 1.38^{\circ}$, an upstream depth of $\dup = 12.5$ cm, and a downstream depth of $\ddn = 3$ cm, which gives the depth ratio $\dratdn = \ddn/\dup= 0.24$. Bolles et al.~systematically varied the driving amplitude in the range $0.125^{\circ}$--$2^{\circ}$ to probe different regimes of wave behavior, from nearly linear waves to strongly nonlinear ones. They found that once a critical amplitude is exceeded (roughly $\Dphi = 0.5^{\circ}$), the downstream statistics become anomalous and are robustly described by the gamma-distribution. Fig.~\ref{ExpSpatialStats} shows spatial statistics for several experiments of different driving amplitudes. The standard deviation of surface displacement $\etastd$ measures the characteristic amplitude of waves (Fig.~\ref{ExpSpatialStats}(a)). Notice that $\etastd$ increases with paddle amplitude, but does not significantly vary in space. In particular, $\etastd$ does not change significantly when waves cross the ADC. Skewness and (excess) kurtosis, however change markedly crossing the ADC once the critical driving amplitude is exceeded. We also show statistics of surface slope $\eta_x$.

 % Figure: Experimental Spatial Statistics
%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{figure}%[!ht]
\begin{center}
\includegraphics[width = 0.9 \linewidth]{Figs/ExpSpatialStats.pdf}
\caption{Wave statistics as they vary in space for several different experiments of different driving amplitudes (see legend). $\etastd$ does not vary significantly crossing the ADC (a), but skewness (b) and kurtosis (c) show a strong response to the ADC. }
\label{ExpSpatialStats}
\end{center}
\end{figure}
 %^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%

>>>>

Rogue waves are abnormally large surface waves, defined by oceanographers as those that exceed twice the significant wave height \citep{muller2005rogue, ying2011linear}. Though such waves were once dismissed as myth, they have now been recorded in oceans across the globe and pose a recognized threat to seagoing vessels and naval structures. Rogue waves, also variously known as freak or anomalous waves, have been observed both in deep and shallow water, and can be triggered by a variety of mechanisms, including anomalous wind-forcing \citep{kharif2008influence, toffoli2017wind}, opposing currents \citep{garrett2009rogue, onorato2011triggering},
focusing due to variable bathymetry \citep{heller2008refraction, white1998chance}, 
and the Benjamin-Feir deep-water modulational instability \citep{benjamin1967disintegration, viotti2013emergence, cousins2015unsteady, farazmand2017reduced}.
The common tie between these mechanisms is their ability to generate non-normal statistics in the surface displacement; when governed by Gaussian statistics, the likelihood of a rogue wave is extremely low, but these rare events occur much more frequently when surface statistics deviate from Gaussian. In this way, it makes sense to approach anomalous waves from the broader perspective of turbulent dynamical systems \cite{sapsis2013a, sapsis2013b, sapsis2013blending, chen2016filtering, majda2016introduction, macedo2017universality, MajdaQiSIAM2018, blonigan2019extreme, guth2019machine}.

@@

Our particular focus is the laboratory experiments of Bolles et al.~\cite{bolles2019}, who demonstrated the emergence of anomalous statistics from a randomized wave-field encountering an abrupt depth change (ADC). In these experiments, the incoming wave-field is generated with nearly Gaussian statistics, and a plexiglass step placed near the middle of the tank creates the depth change. Upon passing into the shallower region, the wave distribution skews strongly towards positive displacement, with deviation from Gaussian being most pronounced a short distance downstream of the ADC. Inspired by these experimental observations, Majda et al.~\cite{majda2019} developed a theoretical framework that accurately captures several key aspects of the anomalous behavior. The theory is based on deterministic and statistical analysis of the truncated Korteweg–de Vries (TKdV) equations, and uses a combination of computational, statistical, and analytical tools. In subsequent work, Majda and Qi \cite{majdaqi2019} analyzed more severely truncated systems that were shown to exhibit many of the same anomalous properties while enjoying a more tractable, lower-dimensional structure.

The purpose of the present manuscript is to provide a more comprehensive treatment of the {\em combined} experimental and theoretical efforts of Bolles et al~\cite{bolles2019} and Majda et al.~\cite{majda2019, majdaqi2019}. Special emphasis is placed on the synergy between experiment and theory -- a cooperative strategy with a proven record of success \cite{camassa2012stratified, ristroph2012, ganedi2018equilibrium}. We present an assortment of previously unpublished experimental data, for example on surface slope statistics and autocorrelation, which motivates further examination and refinement of the theory. Likewise, novel theoretical predictions have inspired us to perform additional experimental measurements, especially as it pertains to a peculiar relationship between surface-displacement skewness and surface-slope variance that is predicted by the theory. Comparison with the new measurements further confirms the predictive power of the framework developed by Majda et al.~\cite{majda2019}. In addition, we provide a more thorough treatment of the link between experimental and model parameters to facilitate comparison between the two. In particular, we flesh out details of non-dimensionalization, which were only briefly addressed in prior work for the sake of brevity.

@@

The pivoting motion of the paddle is driven by a 5-phase stepper motor. To generate a randomized incoming wave field, the paddle angle $\phi$ is specified by a precomputed psuedo-random signal
\begin{align}
\label{PaddleAngle}
& \phi(t) = \phi_0 + \Dphi \sum_{n=1}^N a_n \cos(\omega_n t+\delta_n) \, , \\
\label{anEq}
& a_n = \sqrt{\frac{2 \Delta \omega}{\pi^{1/2} \omsig}} \, 
\exp \left( -\frac{(\omega_n - \omavg)^2}{2 \omsig^2} \right) \, ,
\end{align}
The angular frequencies are evenly spaced $\omega_n = n  \Delta \omega$ with step size $ \Delta \omega = (\omavg+4 \omsig)/N$, where $\omavg$ and $\omsig$ represent the mean and the bandwidth of $\omega$ respectively. As in prior work, all experiments reported here use the values $\omavg = \omsig = 12.5$ rad/s, corresponding to a peak forcing frequency of 2 Hz and bandwidth of 2 Hz. The phases $\delta_n$ are uniformly distributed random variables, which results in a randomized wave train. The standard deviation of the paddle angle, $\Dphi$, controls the amplitude overall amplitude of the waves. In a single experiment $\Dphi$ is fixed, and we will present a series of experiments with $\Dphi$ varied systematically.

@@

	The free surface is illuminated by light-emitting diodes and is imaged from the sideview with a Nikon D3300 at 60 frames per second. The illumination technique, coupled with high pixel count of the camera, allows surface displacements to be resolved with accuracy better than 1/3 millimeter. Furthermore, these optical measurements permit extraction of wave statistics {\em continuously} in space, rather than at a few discrete locations, which is crucial for identifying regions of anomalous wave activity. Further details of the experimental setup can be found in \cite{bolles2019}.

	Example measurements of free-surface displacements $\eta$ are shown in Figs.~\ref{ExpDiagStats}(b)--(c). These measurements are extracted from the images at two representative locations: one a short distance (9 cm) upstream of the ADC and the other a short distance (15 cm) downstream. Both signals exhibit a combination of periodic and random behavior, with the dominant oscillations corresponding to the peak forcing frequency of 2 Hz. The nature of the random fluctuations is revealed by the corresponding histograms shown in Fig.~\ref{ExpDiagStats}(d)--(e). The upstream measurements are symmetrically distributed about the mean, $\eta = 0$. In fact, Bolles et al.~found that these measurements follow a Gaussian distribution closely \cite{bolles2019}. The downstream measurements, however, skew strongly towards positive displacement, $\eta > 0$. Bolles et al.~found these measurements to be well described by a mean-zero gamma distribution \cite{bolles2019}. The slower decay rate of this downstream distribution indicates an elevated level of extreme surface displacement, i.e.~rogue waves. Bolles et al.~estimated that a rogue wave can be up to 65 times more likely in these experiments than if displacements were Gaussian \cite{bolles2019}.

	The paddle amplitude in Fig.~\ref{ExpDiagStats} is $\Dphi = 1.38^{\circ}$, and this value was varied systematically in the range $\Dphi = 0.125^{\circ}$--$2^{\circ}$ to probe the different regimes of wave behavior, from linear to strongly nonlinear waves. Figure \ref{ExpSpatialStats} shows spatial statistics of both surface displacement $\eta$ and the surface slope $\eta_x$ for six different driving amplitudes (see legend). For each quantity, we show the standard deviation, the skewness, and the (excess) kurtosis as they vary in the vicinity of the ADC, located at $x=0$. First, the standard deviation of displacement, $\etastd$, gives the coarsest possible estimate for the amplitude of waves. Figure \ref{ExpSpatialStats}(a) shows that while $\etastd$ increases with driving amplitude, it remains relatively uniform in space for each individual experiment, indicating that the overall amplitude of the wave train is not significantly altered by the presence of the ADC. In experiments with sufficiently high driving amplitude, though, the skewness and kurtosis (b)--(c) respond strongly to the ADC. In these experiments, both the skewness and kurtosis are relatively small upstream of the ADC, indicating nearly Gaussian statistics, but then increase dramatically downstream and reach a peak near $x = 15$ cm. Remarkably, the location of the peak is the same for skewness and kurtosis and appears insensitive to driving amplitude. The maximum values of skewness and kurtosis (roughly 0.9 and 0.7 respectively) seen in the figure indicate a significant departure from Gaussian statistics, as is constant with the histogram in Fig.~\ref{ExpDiagStats}(e). Bolles et al.~found that, once a threshold driving amplitude is exceeded (roughly $\Dphi = 0.5^{\circ}$), the displacement statistics in this anomalous region are robustly described by the gamma distribution.

	The right column of Fig.~\ref{ExpSpatialStats}

 % Figure: Experimental Spatial Statistics
%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{figure}%[!ht]
\begin{center}
\includegraphics[width = 0.8 \linewidth]{Figs/ExpSpatialStats.pdf}
\caption{Wave statistics as they vary in space for several different experiments of different driving amplitudes (see legend). $\etastd$ does not vary significantly crossing the ADC (a), but skewness (b) and kurtosis (c) show a strong response to the ADC. }
\label{ExpSpatialStats}
\end{center}
\end{figure}
 %^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%

\vsp{5}
%The measurements shown in Fig.~\ref{ExpDiagStats} correspond to a paddle driving amplitude of $\Dphi = 1.38^{\circ}$, an upstream depth of $\dup = 12.5$ cm, and a downstream depth of $\ddn = 3$ cm, giving a depth ratio of $\dratdn = \ddn/\dup= 0.24$.
